Estimation of Ne using INLA from a single genealogy
========================================================

This tutorial implements the function in 
http://www.auai.org/uai2012/papers/310.pdf

Example
==================================
We will first simulate two genealogies with 50 tips from a bottleneck demographic scenario. The first genealogy is isochronous and the second one is heterochronous

```{r}
source("simulate_trees.R")
library(ape)

bottleneck_traj<-function(t){
  result=rep(0,length(t))
  result[t<=0.5]<-1
  result[t>0.5 & t<1]<-.1
  result[t>=1]<-1
  return(result)
}

my.out3<-simulate.tree(n=50,N=1,simulator="thinning",Ne=bottleneck_traj,max=11)
my.out3

plot(my.out3$out[[1]],show.tip.label=FALSE)
axisPhylo()
set.seed(123)
samp_times = c(0, sort(runif(40, 0, .8)))
n_sampled = c(10, rep(1, 40))
sample<-cbind(n_sampled,samp_times)
my.out4<-simulate.tree(n=10,N=1,simulator="thinning",Ne=bottleneck_traj,max=11,sampling="hetero",sample=sample)
my.out4

plot(my.out4$out[[1]],show.tip.label=FALSE)

```
For the first genealogy, the following function uses the INLA method to infer Ne at 100 points:
```{r}

source("INLA_Approximation.R")

ci<-coalescent.intervals(my.out3$out[[1]])
data1<-cbind(ci$interval.length,ci$lineages)
result<-.calculate.moller(data1,100,sum(ci$interval.length))


```
Let's compare it with Skyline plot

```{r}
plot(skyline(my.out3$out[[1]]))
points(seq(0,1,.01),bottleneck_traj(seq(0,1,.01)),type="S",lty=2,col="red")


```

