Estimation of Ne using INLA from a single genealogy
========================================================

This tutorial implements the function in 
http://www.auai.org/uai2012/papers/310.pdf

Example
==================================
We will first simulate two genealogies with 50 tips from a bottleneck demographic scenario. The first genealogy is isochronous and the second one is heterochronous

```{r, echo=FALSE}
install.packages("devtools")
library("devtools")
install_github("georgeshirreff/multiNe")
library(multiNe)
library(ape)
#setwd("~/Documents/Hackathon/multiNe/R")
```


```{r}
#install.packages("INLA", repos="http://www.math.ntnu.no/inla/R/stable")

bottleneck_traj<-function(t){
  result=rep(0,length(t))
  result[t<=0.5]<-1
  result[t>0.5 & t<1]<-.1
  result[t>=1]<-1
  return(result)
}

my.out3<-simulate.tree(n=50,N=10,simulator="thinning",Ne=bottleneck_traj,max=11)
my.out3

plot(my.out3$out[[1]],show.tip.label=FALSE)
axisPhylo()
set.seed(123)
samp_times = c(0, sort(runif(40, 0, .8)))
n_sampled = c(10, rep(1, 40))
sample<-cbind(n_sampled,samp_times)
my.out4<-simulate.tree(n=10,N=1,simulator="thinning",Ne=bottleneck_traj,max=11,sampling="hetero",sample=sample)
my.out4

plot(my.out4$out[[1]],show.tip.label=FALSE)

```
For the first genealogy, the following function uses the INLA method to infer Ne at 100 points:
```{r}

library("INLA")
result<-calculate.moller(my.out3$out[[1]],100,1)
par(mfrow=c(1,1))
plot_INLA2(result,traj=bottleneck_traj,main="Bottleneck using INLA")

result2<-calculate.moller(my.out3$out,100,1)
plot_INLA3(result2,traj=bottleneck_traj)

```
Let's compare it with Skyline plot

```{r}
par(mfrow=c(1,2))
plot(skyline(my.out3$out[[1]]))
plot_INLA2(result,traj=bottleneck_traj,main="Bottleneck using INLA")


```

