Simulation of  Genealogies
========================================================

Ape, pegas and coalescentMCMC have functions that simulates genealogies under the coalescent model for specific demographic scenarios. We will consider the following demographic models

The demographic models are

1. Constant Population Size --coalescentMCMC, ape, pegas  

2. Exponential Growth --coalescentMCMC

3. Step Model with one change point--coalescentMCMC

4. Exponential double growth model --coalescentMCMC

5. Any R function for defining $N_{e}$ ---Julia


Simulation of Sequence Data
========================================================
PhyloSim (ape)
Phangorm (simSeq)
phyclust Seq-Gen






Inference from a single genealogy
========================================================
Skyline Plot (ape)
Bayesian Skyline
Bayesian Skyride
Bayesian Skygrid
Bayesian Skytrack
Likelihood Estimator


Inference from sequence data
========================================================





Inference from microsatelite data
========================================================








Inference from Sequence data




Inference from microsatelite data




References: http://cran.r-project.org/web/packages/coalescentMCMC/coalescentMCMC.pdf
ftp://cran.r-project.org/pub/R/web/packages/coalescentMCMC/vignettes/Running_coalescentMCMC.pdf
http://cran.r-project.org/web/packages/coalescentMCMC/vignettes/CoalescentModels.pdf



```{r}
library("ape")
rcoal(5)
plot(rcoal(5))
```

You can also embed plots, for example:

```{r fig.width=7, fig.height=6}
plot(cars)
```

