---
title: "SkylinePlottingExample"
output: html_document
---

This demonstrates how to plot a skyline from a multiPhylo object.

Start by simulating a multiPhylo object from the "ape" package.


```{r}
library(ape)
trs<-rmtree(N=5,n=20)
```

Now if you call the main function this will provide a skyline plot and a table with the median and 2.5 and 97.5 %iles at each time point. 

```{r}
conf.int<-Phylos2Skylines_anchor(trees=trs,output_type="conf.int.plot")
```

The conf.int table can also be plotted directly using conf.int.skyline, with the epsilon value indicating the amount of smoothing (i.e. it specifies the )

```{r}
conf.int.skyline(conf.int,epsilon=0.01)
```

This can also be used to print piecewise linear functions such as the output of INLA.

```{r}
library(multiNe)
tree<-simulate.tree(n=20,N=1)$out[[1]]

ci<-coalescent.intervals(tree)
data1<-cbind(ci$interval.length,ci$lineages)
result<-.calculate.moller(data1,100,sum(ci$interval.length))

conf.int.skyline(result,epsilon=0.01,plot_type="linear")
```
