---
title: "multiNe"
output: html_document
---

Methods to estimate effective population size

```{r, echo=F}
library(ape)
library(multiNe)
library(INLA)
library(adegenet)
```


Start by simulating a multiPhylo object from the "ape" package.
```{r}
trs<-rmtree(N=20,n=100)
```

Now if you call the main function this will provide a skyline plot and a table with the median and 2.5 and 97.5 %iles at each time point. 

```{r}
conf.int<-Phylos2Skylines_anchor(trees=trs,output_type="conf.int.plot")
```

The conf.int table can also be plotted directly using conf.int.skyline, with the epsilon value indicating the amount of smoothing (i.e. it specifies the )

```{r}
conf.int.skyline(conf.int,epsilon=0.01)
```

There is a function for simulating trees under various functions of population size and other factors. 

```{r}
growing_trees<-simulate.tree(n=100,N=20,sampling="hetero")$out
plot(growing_trees[[1]])
conf.int<-Phylos2Skylines_anchor(trees=growing_trees,output_type="conf.int.plot",epsilon=0)
```

This can also be used to graph piecewise linear functions such as the output of INLA.

```{r,eval=T,echo=T}
tree<-rmtree(N=2,n=50)[[1]]
result<-calculate.moller(tree,lengthout=100,L=1)
conf.int.skyline(result,epsilon=0.01,plot_type="linear")
```



```{r,eval=F,echo=F}
library(phyclust)
sim_ms<-ms(nsam=5,nreps=10,"-T")
sim_genind<-ms2ms(sim_ms)
```


```{r}
data(nancycats)
HENe(nancycats)
```
